Linking microbial communities to ecosystem functions: what we can learn from genotype-phenotype mapping in organisms

### 2019-07-09 manuscript/main.tex
Update text with latest revision

\title{Linking microbial communities to ecosystem functions: what we can learn
from genotype-phenotype mapping in organisms} \author{Andrew H. Morris, Brendan
J. M. Bohannan}

@@

@@

Microorganisms mediate many biogeochemical processes and yet the link between
microbial biodiversity and ecosystem function remains unknown. 
Ecologists have investigated the effects of changing
biodiversity on ecosystem function for decades, but this research has primarily
focused on linking terrestrial plant community richness with productivity
\citep{hooper2005}. More recently, there has been an interest in linking
microbial community composition to ecosystem functions such as host disease 
and soil greenhouse gas emissions \citep{schimel1998, singh2010}. 
However, it remains unclear to what extent 
microbial community composition is important for determining the rates of ecosystem-scale
functions.  This uncertainty limits our ability to predict and manage host
health and global biogeochemical cycles. To
address this challenge, we propose integrating traditional
biodiversity-ecosystem function research with ideas from genotype-phenotype
mappingin population genetics. This connection will help determine
whether microbial community
composition alters ecosystem-scale functions independent
of the underlying environmental variation.

If local selection always optimizes the available biodiversity to maximize
ecosystem function, them microbial community composition should not matter for
pedicting the rates of microbially mediated functions. Instead, ecosystem function
should be predicted from the underlying environmental variation. In this case,
the microbial community simply acts as a conduit through which the abiotic
environment alters ecosystem function. However, experiments that manipulate the
connection between environmental factors and microbial community composition
such as environmental treatments, common gardens, and reciprocal transplants
have shown different rates of ecosystem functions for different microbial
communities. This has been observed for plant decomposition
\citep{strickland2009}, plant phenology \citep{panke-buisse2015}, soil nitrogen
cycling \citep{balser2005}, and soil greenhouse gas emissions
\citep{cavigelli2000}. Therefore, variation in microbial community composition
seems to generate variation in ecosystem function. However, in order to
incorporate microorganisms into ecosystem models, we need to identify what
aspects of microbial biodiversity are associated with variation in ecosystem
function.

Most studies of microbial community function focus on one of two aspects of
microbial community structure that are hypothesized to be predictive of
ecosystem function. The first aspect is functional gene or transcript abundance.
In this case, qPCR or shotgun metagonemic sequencing is used to target a gene
or transcript that is a putative marker for a microbial process. For example,
the gene \textit{mcrA}, which encodes the enzyme that performs the final step in
methanogenesis, is hypothesized to be correlated with the rate of
methane production. Other examples include \textit{pmoA} and methanotrophy,
\textit{nifH} and nitrification, and \textit{nosZ} and denitrification. The
rational for this hypothesis is that the abundance of a functional group (as
inferred from marker gene abundance) should correspond to the rate of the
process performed by that group. There are examples where this relationship
holds \citep{freitag2009, freitag2010, schnyder2018}. However, a review of these
studies found that the abundance of a functional gene or transcript is rarely
correlated with the rate of the corresponding process, with most effects either
negative or not significant \citep{rocca2015}.

The second aspect of microbial community structure hypothesized to predict
ecosystem function is taxonomic or functional
diversity. Diversity is either estimated using a measure of community
composition such as 16S rRNA gene ribotyping or manipulated through some proxy
of diversity such as sequential dilution or varying filter sizes. Studies that
experimentally manipulate diversity generally find a relationship between
the applied diversity treatment and ecosystem functions including methanotrophy
\citep{schnyder2018}, phosphorus leaching, greenhouse gas emissions
\citep{wagg2014}, and decomposition \citep{maron2018}. However, in each of these
cases, richness is confounded with other factors such as presence or absence of
major phylogenetic groups (nematodes, fungi) or abundance of microbial cells. In
addition, assembled microbial communities pose the same challenges as
macroorganismal diversity experiments, for example no replication at the highest
diversity treatment. A review of microbial diversity studies shows that
microbial community taxonomic and functional diversity add little variance
explained to models of ecosystem function. Overall, functional gene and
abundance and community doversity improve models of ecosystem function less than
one third of the time and increase variance explained by only 8 percentage
points \citep{graham2016}.

These approaches follow on the tradition of macroorganismal
biodiversity-ecosystem function research by measuring or manipulating diversity
and comparing it to function or by looking at the association between richness
or abundance of a functional group (e.g. plants) and the function performed by
that group (e.g. productivity). However, in the case of microbes, our knowledge
of functional groups is quite limited. We cannot be sure that we're accurately
estimating the abundance or diversity of soil microbial functional groups the
way we can, say, plants in a grassland.

Organisms exhibit population differentiation due to geographic and ecological
processes \citep{wright1943}. Neutral processes such as local genetic drift and
reduced gene flow due to geographic separation can lead to population
differentiation. In addition, environmental distance can lead to selection
against migrants \citep{hendry2004} and local selection on endemic populations
to generate variation. Microorganisms exhibit biogeographic patterns due to
ecological selection and geographic separation \citep{martiny2006}. If dispersal
is non-limiting then the most productive species should always be present and
composition should not matter for local ecosystem function. However, if important
functional groups are dispersal-limited, lost to drift, or are unable to
colonize a site, then there could be important functional consequences of
community differentiation to geographic or ecological isolation.

There are a numbber of ways that environmental or spatial distance could obscure
microbial BEF relationships. For example, historical effects of geographic
isolation such as dispersal limitation could lead to spurious associations
between the abundance of taxa and the rate of ecosystem function. Community
stratification where taxa co-occur due to historical effects or local selection.
Environmental selection can also shape microbial community
structure. For example, pH is known to regulate both broad biodiversity patterns for
microbial communities \citep{fierer2006} and methane cycling dynamics
\citep{ye2012}. This could like to microbial taxa that are correlated with
methane cycling, but are mechanisticalyl unrelated. In fact, if the rate of
ecosystem function is high in one set of ecosystems, function will be
associated with any taxa that are abundant in that set of
ecosystems.

In order to understand the biological basis of ecosystem function, we need to
separate the close association between environmental variation, community
composition, and ecosystem function in order to identify what aspects of
microbial biodiversity determine the rate of ecosystem function.

The challenge of mapping microbial community structure to ecosystem function is
analogous to the challenge of identifying the genetic basis of complex traits in
organismal populations. In the latter case, a population exhibits variation in a
phenotype (e.g. hight or disease state) as well as variation in potentially
thousands of genetic loci (usually identified as single nucleotide polymorphisms
(SNPs)). To identify the genetic basis for a trait, investigators sample from
this population and correlate phenotype with genotype. In most cases, there are
many more loci than individuals and we do not know the function of genes
associated with these SNPs. Genetic markers identified in these genome-wide association
studies can be the target for future research to dissect the function of those
genes, can be added or removed to test their association with the trait, or can
be used as marker for that disease or phenotype even if it has no known
mechanistic association.

This mirrors the problem of microbial
structure-function mapping. In this case, a population of hosts or ecosystems exhibit
variation in a particular trait or ecosystem function putatively mediated by
microorganisms as well as variation in metagenomic composition. Investigators
could sample from this population of ecosystems and correlate the rate of an
ecosystem function with microbial community composition or gene composition.

Population geneticists trying to make robust associations between genetic loci
and phenotype must take into account population stratification. Population
stratification due to shared ancestry or local selection can lead to spurious
associations between genes and traits. For example, human height varies
across Europe in a latitudinal gradient. Positive associations between genes and
phenotypes can arise because of true associations, through linkage
disequilibrium with true genes, or through admixture \citep{lander1994}.
Microbial structure-function connections could pose these same problems. There
could be true associations between a microorganism and the ecosystem function
performed by that taxon, or there could be there could be systematic
co-occurrence patterns between one taxon and a causal taxon like metabolic
networks, or there could be admixture due to shared history or selection within
a sub-set of ecosystems. Robust structure-function associations rely on
controlling for these factors.

>>>>

\title{Linking microbial communities to ecosystem functions: what we can learn
from genotype-phenotype mapping in organisms}

\author{Andrew H. Morris, Brendan J. M. Bohannan}

@@

\section*{Abstract}

@@

Microorganisms mediate many biogeochemical processes and yet the link between
microbial biodiversity and ecosystem function remains unknown. 
Ecologists have investigated the effects of changing
biodiversity on ecosystem function for decades, but this research has primarily
focused on linking terrestrial plant community richness with productivity
\citep{hooper2005}. More recently, there has been an interest in linking
microbial community composition to ecosystem functions such as host disease 
and soil greenhouse gas emissions \citep{schimel1998, singh2010}. 
However, it remains unclear to what extent 
microbial community composition is important for determining the rates of ecosystem-scale
functions.  This uncertainty limits our ability to predict and manage host
health and global biogeochemical cycles. To
address this challenge, we propose integrating traditional
biodiversity-ecosystem function research with ideas from genotype-phenotype
mapping in population genetics. This connection will help determine
whether microbial community
composition alters ecosystem-scale functions independent
of the underlying environmental variation.

If local selection always optimizes the available biodiversity to maximize
ecosystem function, them microbial community composition should not matter for
pedicting the rates of microbially-mediated functions. Instead, 
the rates of these functions should be determined by the
underlying environmental variation. In this case, the microbial community simply
acts as a conduit through which the abiotic
environment alters ecosystem function. However, experiments that manipulate the
connection between environmental factors and microbial community composition
such as environmental treatments, common gardens, and reciprocal transplants
have shown different rates of ecosystem functions for different microbial
communities. This has been observed for plant decomposition
\citep{strickland2009}, plant phenology \citep{panke-buisse2015}, soil nitrogen
cycling \citep{balser2005}, and soil greenhouse gas emissions
\citep{cavigelli2000}. Therefore, variation in microbial community composition
seems to generate variation in ecosystem function. However, in order to
incorporate microorganisms into ecosystem models, we need to identify what
aspects of microbial biodiversity are associated with variation in ecosystem
function.

Most studies of microbial community function focus on one of two aspects of
microbial community structure that are hypothesized to be predictive of
ecosystem function. The first aspect is functional gene or transcript abundance.
In this case, qPCR or shotgun metagonemic sequencing is used to target a gene
or transcript that is a putative marker for a microbial process. For example,
the gene \textit{mcrA}, which encodes the enzyme that performs the final step in
methanogenesis, is hypothesized to be correlated with the rate of
methane production. Other examples include \textit{pmoA} and methanotrophy,
\textit{nifH} and nitrification, and \textit{nosZ} and denitrification. The
rationale for this hypothesis is that the abundance of a functional group (as
inferred from marker gene abundance) should correspond to the rate of the
process performed by that group. There are examples where this relationship
holds \citep{freitag2009, freitag2010, schnyder2018}. However, a review of these
studies found that the abundance of a functional gene or transcript is rarely
correlated with the rate of the corresponding process, with most effects either
negative or not significant \citep{rocca2015}.

The second aspect of microbial community structure hypothesized to predict
ecosystem function is taxonomic or functional
diversity. Diversity is either estimated using a measure of community
composition such as 16S rRNA gene ribotyping or manipulated through some proxy
of diversity such as sequential dilution or varying filter sizes. Studies that
experimentally manipulate diversity generally find a relationship between
the applied diversity treatment and ecosystem functions including methanotrophy
\citep{schnyder2018}, phosphorus leaching, greenhouse gas emissions
\citep{wagg2014}, and decomposition \citep{maron2018}. However, in each of these
cases, richness is confounded with other factors such as presence or absence of
major phylogenetic groups (nematodes, fungi) or abundance of microbial cells. In
addition, assembled microbial communities pose the same challenges as
macroorganismal diversity experiments, for example no replication at the highest
diversity treatment. A review of microbial diversity studies shows that
microbial community taxonomic and functional diversity add little variance
explained to models of ecosystem function. Overall, functional gene
abundance and community diversity improve models of ecosystem function less than
one third of the time and increase variance explained by an average of only 8 percentage
points \citep{graham2016}.

These approaches follow on the tradition of macroorganismal
biodiversity-ecosystem function research by measuring or manipulating diversity
and comparing it to function or by looking at the association between richness
or abundance of a functional group (e.g. plants) and the function performed by
that group (e.g. productivity). However, in the case of microbes, our knowledge
of functional groups is quite limited. We cannot be sure that we're accurately
estimating the abundance or diversity of soil microbial functional groups the
way we can plants in a grassland.

Organisms exhibit population differentiation due to geographic and ecological
processes \citep{wright1943}. Neutral processes such as local genetic drift and
reduced gene flow due to geographic separation can lead to population
differentiation. In addition, environmental distance can lead to selection
against migrants \citep{hendry2004} and local selection on endemic populations
to generate variation. Microorganisms also exhibit biogeographic patterns due to
ecological selection and geographic separation \citep{martiny2006}. If dispersal
is non-limiting then the most productive species should always be present and
composition should not matter for local ecosystem function. However, if important
functional groups are dispersal-limited, lost to drift, or are unable to
colonize a site, then there could be important functional consequences of
community differentiation to geographic or ecological isolation.

There are a number of ways that environmental or spatial distance could obscure
microbial BEF relationships. For example, historical effects of geographic
isolation such as dispersal limitation could lead to spurious associations
between the abundance of taxa and the rate of ecosystem function. Community
stratification where taxa co-occur due to historical effects or local selection.
Environmental selection can also shape microbial community
structure. For example, pH is known to regulate both broad biodiversity patterns for
microbial communities \citep{fierer2006} and methane cycling dynamics
\citep{ye2012}. This could lead to microbial taxa that are correlated with
methane cycling, but are mechanistically unrelated. In fact, if the rate of
ecosystem function is high in one set of ecosystems, function will be
associated with any taxa that are abundant in those ecosystems. 
In order to understand the biological basis of ecosystem function, we need to
separate the close association between environmental variation, community
composition, and ecosystem function in order to identify what aspects of
microbial biodiversity determine the rate of ecosystem function.

The challenge of mapping microbial community structure to ecosystem function is
analogous to the challenge of identifying the genetic basis of complex traits in
organismal populations. In the latter case, a population exhibits variation in a
phenotype (e.g. height or disease state) as well as variation in potentially
thousands of genetic loci (usually identified as single nucleotide polymorphisms
(SNPs)). To identify the genetic basis for a trait, investigators sample from
this population and correlate phenotype with genotype. In most cases, there are
many more loci than individuals and we do not know the function of genes
associated with these SNPs. Genetic markers identified in these genome-wide association
studies can be the target for future research to dissect the function of those
genes, can be added or removed to test their association with the trait, or can
be used as marker for that disease or phenotype even if it has no known
mechanistic association. This mirrors the problem of microbial
structure-function mapping. In this case, a population of hosts or ecosystems exhibit
variation in a particular trait or ecosystem function putatively mediated by
microorganisms as well as variation in metagenomic composition. Investigators
could sample from this population of ecosystems and correlate the rate of an
ecosystem function with microbial community composition or gene composition.

Population geneticists trying to make robust associations between genetic loci
and phenotype must take into account population stratification. Population
stratification due to shared ancestry or local selection can lead to spurious
associations between genes and traits. For example, human height varies
across Europe in a latitudinal gradient. Positive associations between genes and
phenotypes can arise because of true associations, through linkage
disequilibrium with true genes, or through admixture \citep{lander1994}.
Microbial structure-function connections could pose these same problems. There
could be true associations between a microorganism and the ecosystem function
performed by that taxon, or there could be systematic
co-occurrence patterns between one taxon and a causal taxon like metabolic
networks, or there could be admixture due to shared history or selection within
a subset of ecosystems. This is true for ASVs in ribotyping 16S datasets and for
genes in metagenomic (or other -omic) datsets. 
Robust structure-function associations rely on
controlling for these factors.

\subsection{Sample collection}

\subsection{Determination of the rates of ecosystem functions}

\subsection{Measurement of environmental variables}

\subsection{Sequencing and bioinformatic processing}

DADA2 pipeline for amplicon sequence variant (ASV) assignment.

We used the variance stabilizing transformation from the \texttt{DESeq2}
package to adjust for differences in sequencing depth between samples.

\subsection{Statistical analysis}

All statistical analyses were performed using the \texttt{R} statistical
environment. 
We first applied traditional tests of microbial biodiversity-ecosystem function
relationships. To do this, we fit ordinary least squares regression models with
the rate of ecosystem function as the response with one predictor for each
model. These predictors include functional gene and transcript abundance,
diversity measures including richness, Shannon's diversity index, and Simpson's
diversity index, measures of microbial community composition using the first
several principal component axes from a principal component
analysis (PCA) chosen based on visual inspection of a screeplot incpluding a PCA for
abundance data and a PCA for presence/absence data, and several 
environmental variables.

Next, we used a statistical approach common in population genetics genome-wide
association studies. We fit variance component mixed models from the
\texttt{varComp} package to test whether variation in microbial
community composition expains a significant portion of the variation in
ecosystem function \citep{qu2013}. First, we fit an intercept-only model with one random effect
for community similarity calculated from Bray-Curtis dissimilarity for abundance
data and the Jaccard index for presence/absance data. We tested the significance
of these components using a linear score test by comparing nested models with and
without the microbial community variance component \citep{qu2013}. We then estimated
the community heritability of this function by calculating the degree of
variation in the rate of ecosystem function due to variation in community
composition between ecosystems including standard errors on that estimate 
using the \texttt{h2G} function from the
\texttt{gap} package. We then tested whether microbial community composition explains a
significant portion of the variation in ecosystem function after accounting for
variation in the environment and accounting for the spatial distance between
samples. We did this by fitting an intercept-only model with variance components
for environmental similarity from euclidean distance of a matrix of
environmental covariates and spatial similarity
also using euclidean distance from the latitude and longitude coordinates where
the samples were collected. We then used a linear score test to compare
nested models with and without the microbial community variance component to
determine whether community composition adds any variance explained after
accounting for spatial and environmental covariates. We then estimated the
variance explained by each component (environment, geography, and community
composition) with standard errors of the estimates using the \texttt{h2GE}
function.

After determining which variance components were significant, we tested the
effect of each ASV on the rate of ecosystem function. We fit a variance
component model for each ASV with ASV abundance as the fixed effect and random
effects for each of the variance components. This tests the relationship between
taxon abundance and the rate of ecosystem function after controlling for
population stratification and data structure. Significant ASVs were identified
by controlling the false discovery rate

\begin{itemize}
  \item For low-affinity and high-affinity methane oxidation in Gabon and
    methane production in Brazil, community composition explains a significant
    portion of the variation in ecosystem function. (linear score test or
    likelihood ratio test)
  \item On its own, the portion of `community heritability' varies by function.
    \textasciitilde99\% (SE = 20\%) for high affinity, \textasciitilde61\% (SE =
    39\%)for low affinity, etc. 
  \item Geographic proximity: high affinity 0.85 (0.10), low affinity n.s. 0.34
    (0.31)
  \item Environmental similarity: high affinity 0.61 (0.23), low affinity 0.61 (0.36) 
  \item Some of these become n.s. when combined suggesting they explain an
    overlapping portion of variation. Perhaps this means we should fit the
    variance component model with just the community similarity matrix knowing
    that that will capture the important components of historical effects and
    selection.
  \item Function varies strongly by ecosystem type: in Brazil, pasture has
    higher methane fluxes than primary or secondary forest; in Gabon, peatlands
    and wetlands have higher low affinity while plantations and forests have
    higher high affinity.
  \item Possible next step: fit a variance component model with community
    similarity as the random effect and taxon abundance as the fixed effect and
    collect p.values for each taxon (though this may return none). Or include
    all three variance components. Adjust for false discovery rate to get
    significant taxa.
\end{itemize}

\section{Discussion}

In this paper, we argue that the traditional approach to microbial BEF research
ignores important understandings about the nature of ecological processes in
determining the distribution of microorganisms. The marker gene hypothesis has
proven unsuccesful in predicting the rates of ecosystem functions from the
abundance of functional genes and in our example study was not significant in one case and
was significant in another, though explained a small amount of the variation.
Similarly, diversity metrics from the entire Prokrayotic community were not
explanatory. These results agree with recent reviews of this literature and
demonstrate that these approaches are not a fruitful avenue for elucidating
microbial structure-function connections.

Instead, microbial ecologists should treat microbial communities as analogous to
organisms with the metagenome representing the community genome. In this case,
ecosystem function represents a phenotype at the community level. As in
organismal biology, this ecosystem phenotype will be a mix of genotypic effects and
environmental effects. Using this conceptual framework, population genetic
approaches can be applied to microbial community datasets to identify

Treating ecosystems like organisms. Calculating community heritability. Possible
approaches: variance component modeling, experimental ecosystem
selection/experimental evolution, CAFE and the Price equation. Need to collect
data in such a way to optimally use these techniques that already exist.

\section{Conclusions}
Microbial community ecosystem function research has demonstrated that
functional gene abundance is not a good predictor of ecosystem function and
that diversity metrics do not add much to ecosystem models. Therefore, we
argue that microbial ecologists should look more agnostically at microbial
community structure to identify whether microbial community composition is
important for determing the rate of ecosystem function across space, time,
and ecological conditions. In this way we can identify which taxa or aspects of
microbial community structure determine the rates of ecosystem functions and so
will be useful for predicting and managing those rates.

The problem of
connecting a complex microbial community matrix with an ecosystem-scale function
is not unique to microbial ecology. Population geneticists attempting to like
organismal phenotypes to its genotypic and environmental basis have dealt with
this problem for decades. In order to make robust associations between the
presence or abundance of specific microbial taxa and the rate of ecosystem-scale
functions, we should take lessons from population genetics in identifying
correlations between community structure and ecosystem function. We suggest
three approaches to addressing this question:

\begin{enumerate}
  \item Metagenome-wide association studies/Community-wide association
    studies/Ecosystem-wide association studies to survey natural ecosystems
    for potential biomarkers or drivers of ecosystem function. Appropriately
    control for community structure patterns to make robust
    structure-function associations. Combine this
    with experimental manipulations such as reciprocal transplants or
    environmental treatments to identify community-environment (CxE)
    interactions.
  \item Use artificial ecosystem selection to enrich for groups of taxa that
    respond to selection on a function. Use this to identify organisms that
    might play a critical role in ecosystem function. Use a response to
    selection to calculate ``community heritability''or the prortion of
    variation in function due to community composition.
  \item Measure rates of ecosystem functions on the individual cell or
    individual taxon level in order to partition changes in ecosystem
    function to gains, losses, or changes in abundance of
    specific taxa using the Price equation. Accomplish this by combining
    isotope source partitioning with functional group assessments, using
    labeling approaches like stable-isotope probing, ...
\end{enumerate}

\section*{Old Text}
